Hot Nock: A History of the Nock Combinator Calculus

### 2025-07-06 mss.tex
Post

Nock is a combinator calculus which serves as the computational specification layer for the Urbit and Nockchain/NockApp systems.  It is a hyper-\textsc{risc} instruction set architecture (ISA) intended for execution by a virtual machine (but see \citet{Mopfel2025}, pp.~XX–XX herein).  Nock's simplicity and unity of expression make it amenable to proof-based reasoning and guarantees of correctness.  Its Lisp-like nature surfaces the ability to introspect on the code itself, a property which higher-level languages compiling to it can exploit.  Nock permits itself a finite number of specification changes, called ``decrements'' or ``kelvins'', which allow it to converge on a balance of expressiveness and efficacy.  This article outlines the extant historical versions of the Nock combinator calculus and reconstructs the motivation for the changes made at each kelvin decrement.  It begins with an exposition of Nock as a tool of computation, outlines the history of Nock's decrements, and speculates on motivations for possible future developments.

@@

    \begin{quote}
    Everything about a scope, including name bindings, aliases, and docstrings, is stored in the subject's type. This allows Hoon's compilation discipline to be similarly minimal: the compiler is a function from subject type and Hoon source to product type and compiled Nock. Running this Nock against a value of the subject type produces a vase of the result. It's hard to imagine a more streamlined formalism for compilation.  \citep{Blackman2020}
    %(Ted Blackman, [“Why Hoon”](https://urbit.org/blog/why-hoon))
    \end{quote}

    \noindent
    Some Nock opcodes alter the subject, such as a variable declaration, by producing a new subject which is utilized for subsequent axis lookups.

@@

We have asserted without demonstration thus far that Nock is a combinator calculus.  We now show that this is the case, with reference to Nock 4K, the latest specification.  The simplest combinator calculus consists of only three combinators \citep{SomeoneSmart}, \textsc{s}, \textsc{k}, and \textsc{i}.  These combinators are defined as follows:

\begin{enumerate}
  \item  \textsc{s} substitution.  This corresponds to Nock 4K's opcode 2, which substitutes the second argument into the first argument at the third argument's axis.  (There are some subtle differences to Nock's expression of \textsc{s} as opcode 2 that we will elide as being fundamentally similar, but perhaps worthy of its own monograph.)
  \item  \textsc{k} constant.  This corresponds to Nock 4K's opcode 1, which yields its argument as a constant noun.
  \item  \textsc{i} identity.  This corresponds to Nock 4K's opcode 0, a generalized axis lookup operator, which can trivially retrieve the current subject or expression as well as any children.
\end{enumerate}

\noindent
While Nock introduces a few more primitive operations as a practicality, the above identities establish its bona fides as a combinator calculus capable of general computation.  Similar to Haskell Curry's \textsc{bckw} system, which can be written in forms reducible to \textsc{ski}, Nock provides a set of primitive rules and a set of economic extended rules for convenience in writing a compiler.  (See \citet{Galebach2025}, pp.~XX–XX in this volume, for exposition on how to evaluate a Nock expression by hand or by interpreter.)

@@

This latter idea he particularly connected to the concept of a ``scry namespace'':  ``dereferencing Urbit paths is as natural (and stateless) a function as increment or equals'' (ibid.).  Indeed, Urbit's current userspace utilizes such an affordance to replicate a global scope environment for accessing system and remote resources.  (See \citet{Davis2025b}, pp.~XX–XX in this volume, for a discussion of the Nock virtualized interpreter.)
%Yarvin2010:  https://github.com/cgyarvin/urbit/blob/1bc9086a3fd62594e8bd631744cb9bf804f6a43b/Spec/urbit/5-whynock.txt

@@

The Nock family survives in a trail of breadcrumbs, with each version of the specification being a kelvin decrement of the previous version.  Early versions were produced exclusively by Curtis Yarvin, eventually involving the input of other developers after the 2013 founding of Tlon Corporation.  In this section, we present each extant version of the Nock specification and comment on the changes and their motivations.  Only the layouts have been changed for print.  Dates for Nock specifications are derived from dated public posts (U, 9K), internal dating (13K, 12K, 11K, 10K), or from Git commit history data (8K, 7K, 6K, 5K).\footnote{In at least one case (7K), Yarvin claims to have finished the proposal a month earlier but to not have posted it until this date.}  No version of 14K survives publicly, nor does any primordial version prior to \texttt{U} (15K) appear to exist.

@@

\subsection{U}

\begin{quote}
I have not really worked with combinator models, but my general impression is that it takes essentially an infinite amount of syntactic sugar to turn them into a programming language. \texttt{U} certainly takes some sweetener, but not, I think, as much.  (Curtis Yarvin)
\end{quote}

The earliest extant Nock is \texttt{U}, a proto-Nock posted to the \textit{Lambda the Ultimate} blog in 2006 \citep{Yarvin2006}.\footnote{Curtis Yarvin was consulted for elements of this history.  Unfortunately many elements of the original prehistory of Nock appear to be lost to the sands of time on unrecoverable hard drives.}  The draft is versioned 0.15; subsequent evidence indicates that this is a downward-counting kelvin-versioned document already.  The specification is produced in full in Listing~\ref{lst:u}.
% Yarvin2006, http://lambda-the-ultimate.org/node/1269
% Yarvin2006a, https://web.archive.org/web/20060701070718/http://urbit.sourceforge.net/u.txt

Extensive commentary on the operators is provided.  Rightwards grouping of tuple expressions has already been introduced.  Extension of the language is summarily ruled out.\footnote{Compare Ax and Conk, pp.~XX–XX herein.}  Details of parsing and lexing are considered.  Terms (the ancestor of nouns) include a \texttt{NULL}-like ``foo'' type \texttt{\textasciitilde} distinguishable by value rather than structure.  \textsc{ascii} is built in as numeric codes, similar to G\"{o}del numbering.

As commenter Mario B.\ pointed out, the U specification permits \textsc{ski} operators with the simple expressions,

@@

While early work (1940s--50s) had been carried out on ``minimal instruction set computers'' (\textsc{misc}s), it is more likely that Yarvin was influenced by contemporaneous work on ``reduced instruction set computers'' (\textsc{risc}s) in the 1980s and 90s.  Language proposals like that of \citet{Burger2006} (Pico Lisp) may have influenced Yarvin's design choices throughout this era.
%Burger2006 https://software-lab.de/radical.pdf

@@

\begin{lstlisting}[label=lst:u,caption={U, 31 January 2006.  The earliest extant patriarch of the Nock family.},style=listingcode]
U: Definition

@@

In plain language, the rules of \texttt{U} are as follows:

\begin{enumerate}[label=(\alph*)]
  %a
  \item  axis lookup
  %b
  \item  cell detection
  %c
  \item  atom detection
  %d
  \item  constant
  %e
  \item
  %f
  \item  equality
  %g
  \item  increment
  %h
  \item
  %i
  \item
  %j
  \item
  %k
  \item
  %l
  \item  cell distribution
  %m
  \item  head
  %n
  \item
  %o
  \item
  %p
  \item  hyper-Turing follow operator
  %q
  \item
  %r
  \item  cell distribution
  %s
  \item  trivial evaluation
\end{enumerate}

@@

Axiomatic operator \texttt{*} is identified as a \texttt{GOTO}.

@@

@@

Opcodes were reordered slightly.  A dynamic hint opcode 8 and a static hint opcode 9 were added.

TODO is that 8?

@@

@@

TODO

@@

@@

Parentheses were replaced by brackets.  Opcodes were reordered slightly.  Compound opcodes for composition and extension (variable definition) were removed.  Functionally, 11K and 10K appear very similar, particularly if the Watt (proto-Hoon) compiler is set up to produce variable declarations and compositions as the compound opcodes had them.

@@

@@

The cell detection axiomatic operator underlying opcode 4 (cell detection) was changed from \texttt{\&} to \texttt{?}.

This version of Nock was published on the Moron Lab blog in 2010 \citep{Yarvin2010} as ``Maxwell's equations of software''.  Yarvin emphasized that Nock was intended to serve as ``foundational system software rather than foundational metamathematics'' (ibid.).

Yarvin also publicly exposed his thoughts about building a practical higher-level language on top of Nock at this point (ibid.):

\begin{quote}
To define a language with Nock, construct two nouns, \texttt{q} and \texttt{r}, such that \texttt{*[q r]} equals \texttt{r}, and \texttt{*[s *[p r]]} is a useful functional language. In this description,
\begin{itemize}
  \item  \texttt{p} is the function source;
  \item  \texttt{q} is your language definition, as source;
  \item  \texttt{r} is your language definition, as data;
  \item  \texttt{s} is the input data.
\end{itemize}
\end{quote}
TODO bug in expression?

More concretely, Watt (the predecessor to Hoon) is defined as:

@@

\noindent
TODO

Versus 10K, 9K elides operator names in favor of definitions.  Other differences are likewise primarily terminological, such as the replacement of \texttt{Deep} \texttt{\&} with \texttt{?}.

Yarvin has been to think about virtualizing Nock interpretation to expose a broader namespace for interaction with values than the ``strict'' subject of a formula \citep{Yarvin2010}.

@@

@@

The compound opcodes reappear for the first time since \texttt{U} in Nock 8K.  An opcode 12 appears as TODO

8K received an uncharacteristic amount of commentary, given a document prepared for its presentation at \textsc{sss10} \citep{Yarvin2010a}.
% Yarvin2010a, https://github.com/cgyarvin/urbit/blob/900ef3b8ffc4a7245b74b816566ad23051a371b1/Spec/watt/sss10.tex

Lambda was highlighted as a design pattern (a ``gate'' or stored procedure call) enabled by the ``core'' convention.  Notably, \lstinline[style=listingcode]{[[sample context] battery]} occurred in a different order than has been conventional since 2013 (emphasizing that the common core pattern is a convention rather than a requirement of Nock).

@@

@@

TODO

@@

Opcode 12 is reduced to TODO

@@

@@

TODO

The axiomatic operator for increment was changed from \texttt{\^} to \texttt{+}.  The definition of opcode 8 (extend) was simplified somewhat, and the order XXX [0 2]

@@

@@

All trivial reductions of axiomatic operators were removed to the preface of the specification.

(For instance, a trivial ``cosmetic'' change was made to 5K's specification after it was publicly posted in order to synchronize it with the VM's behavior
% https://github.com/urbit/archaeology/commit/dd779c1e92d7200c2ce6c922052f1eb1d1642e4c
)

@@

@@

The primary change motivating 5K to 4K was the introduction of an edit operator \texttt{#}, which ameliorated the proliferation of cells in the Nock runtime's memory.  The edit operator is an optimization which makes modifications to a Nock data structure more efficient.  It's an interesting example of a change motivated by the runtime rather than theoretical or higher-level language concerns.

Opcode 5 (equality) was rewritten to more explicit with application of the cell distribution rule.  Opcodes 6–9 were rewritten to utilize the \texttt{*} operator rather than routing via opcode 2.  Opcode 11 (formerly opcode 10) was likewise massaged.  In general, preferring to express rules using \texttt{*} proved to be slightly more terse than utilizing opcode 2.

\begin{lstlisting}[label=lst:nock4k,caption={Nock 4K, October 2018.},style=listingcode]
Nock 4K

@@

@@

While deviations from the trunk line of the Nock family have been propsed at various points,\footnote{Notably, Ax/Conk (see pp.~XX–XX, this volume), \textsc{skew}, and \textsc{plan} (see pp.~XX–XX, this volume).} Nock itself has remained the definitional substrate of Urbit since its inception.  It has also been adopted as the primary \textsc{isa} of Nockchain and the NockApp ecosystem.

Why, then, do we contemplate further changes?  The \textsc{skew} proposal by \patp{siprel} and \patp{little-ponnys} argued that Nock 4K represented an undesirable saddle point in the design space of possible Nocks, itself a ``ball of mud'' \citep{Skew2020}.  While \textsc{skew} itself was not adopted, it inspired the development of \textsc{plan} as a solid-state computing architecture sharing some ambitions with Urbit and Nock.  A rigorously æsthetic argument can thus be sustained that Nock is not yet ``close enough'' to its final, diamond-perfect form to be a viable candidate.

While some have found this argument compelling, Urbit's core developers have elected to maintain work in the ``main line'' of Nock as an \textsc{isa}.  The Nock 4K specification is a good candidate for a ``final'' version of Nock, as it is the most recent version and has been used in production for several years.  It seems more likely that subsequent changes to Nock will derive not from alternative representations but from either dramatically more elegant expressions (e.g., of opcode 6) or from an implicit underspecification in the current Nock 4K which should be made explicit.

@@

Nock began life as a hyper-Turing machine language, a theoretical construct for the purpose of defining higher-level programming languages with appropriate affordances and semantics.  While its opcodes and syntax have gradually evolved over the course of two decades, the ambition to uproot the Unix ``ball of mud'' and replace it with a simple operating function amenable to reason has remained the north star of Urbit and Nock.

>>>>

Nock is a combinator calculus which serves as the computational specification layer for the Urbit and Nockchain/NockApp systems.  It is a hyper-\textsc{risc} instruction set architecture (ISA) intended for execution by a virtual machine (but see \citet{Mopfel2025}, pp.~XX–XX herein).  Nock's simplicity and unity of expression make it amenable to proof-based reasoning and guarantees of correctness.  Its Lisp-like nature surfaces the ability to introspect on the code itself, a property which higher-level languages compiling to it can exploit.  Yet for all this, Nock was not born from a purely mathematical approach, but found its roots in practical systems engineering.

Nock permits itself a finite number of specification changes, called ``decrements'' or ``kelvins'', which allow it to converge on a balance of expressiveness and efficacy.  This article outlines the extant historical versions of the Nock combinator calculus and reconstructs the motivation for the changes made at each kelvin decrement.  It begins with an exposition of Nock as a tool of computation, outlines the history of Nock's decrements, and speculates on motivations for possible future developments.

@@

  Some Nock opcodes alter the subject (for instance a variable declaration) by producing a new subject which is utilized for subsequent axis lookups.

@@

We have asserted without demonstration thus far that Nock is a combinator calculus.  We now show that this is the case, with reference to Nock 4K, the latest specification.  The simplest combinator calculus consists of only three combinators: \textsc{s}, \textsc{k}, and \textsc{i} \citep{Wolfram2021}.  These combinators are:

\begin{enumerate}
  \item  \textsc{s} substitution.  $\textsc{s}xyz = xz(yz)$, returns the first argument applied to the third, then applies this to the result of the second argument applied to the third.  This corresponds to Nock 4K's opcode 2, which substitutes the second argument into the first argument at the third argument's axis.  (There are some subtle differences to Nock's expression of \textsc{s} as opcode 2 that we will elide as being fundamentally similar, but perhaps worthy of its own monograph.)
  \item  \textsc{k} constant.  $\textsc{k}xy = x$, consumes its argument and returns a constant in all cases.  This corresponds to Nock 4K's opcode 1, which yields its argument as a constant noun.
  \item  \textsc{i} identity.  $\textsc{i}x = x$, returns its argument.  This corresponds to a special case of Nock 4K's opcode 0, a generalized axis lookup operator, which can trivially retrieve the current subject or expression as well as any children.
\end{enumerate}
%Wolfram2021 https://writings.stephenwolfram.com/2020/12/combinators-a-centennial-view/ but the book

\noindent
While Nock introduces a few more primitive operations as a practicality, the above identities establish its bona fides as a combinator calculus capable of general computation.  Similar to Haskell Curry's \textsc{bckw} system, which can be written in forms isomorphic to \textsc{ski}, Nock provides a set of primitive rules and a set of economic extended rules for convenience in writing a compiler.\footnote{See \citet{Galebach2025}, pp.~1–45 in this volume, for exposition on how to evaluate a Nock expression by hand or by interpreter.}

@@

This latter idea he particularly connected to the concept of what came to be called a ``scry namespace'':  ``dereferencing Urbit paths is as natural (and stateless) a function as increment or equals'' (ibid.).  Indeed, Urbit's current userspace utilizes such an affordance to replicate a global scope environment for accessing system and remote resources.  (See \citet{Davis2025b}, pp.~XX–XX in this volume, for a discussion of the Nock virtualized interpreter.)
%Yarvin2010:  https://github.com/cgyarvin/urbit/blob/1bc9086a3fd62594e8bd631744cb9bf804f6a43b/Spec/urbit/5-whynock.txt

@@

The Nock family survives in a trail of breadcrumbs, with each version of the specification being a decrement of the previous version.\footnote{This system, called ``kelvin decrementing'', draws on analogy with absolute zero as the lowest possible temperature---and thus most stable state.}  Early versions were produced exclusively by Curtis Yarvin, eventually involving the input of other developers after the 2013 founding of Tlon Corporation.  In this section, we present each extant version of the Nock specification and comment on the changes and their motivations.  Only the layouts have been changed for print.  Dates for Nock specifications were derived from dated public posts (\texttt{U}, 9K), internal dating (13K, 12K, 11K, 10K), or from Git commit history data (8K, 7K, 6K, 5K).\footnote{In at least one case (7K), Yarvin claims to have finished the proposal a month earlier but to not have posted it until this date.}  No version of 14K survives publicly, nor does any primordial version prior to \texttt{U} (15K) appear to exist.

@@

\subsection[\texttt{U}]{U}

\begin{quote}
I have not really worked with combinator models, but my general impression is that it takes essentially an infinite amount of syntactic sugar to turn them into a programming language. \texttt{U} certainly takes some sweetener, but not, I think, as much.  (Curtis Yarvin, 2006)
\end{quote}

\sloppy
The earliest extant Nock is \texttt{U}, a proto-Nock posted to the \textit{Lambda the Ultimate} blog in 2006 (\citet{Yarvin2006a}; \citet{Yarvin2006}).\footnote{Curtis Yarvin was consulted for elements of this history.  Unfortunately many elements of the original prehistory of Nock appear to be lost to the sands of time on unrecoverable hard drives.}  The draft is versioned 0.15; subsequent evidence indicates that this is a downward-counting kelvin-versioned document already.  The full specification is reproduced in Listing~\ref{lst:u}.\footnote{The original text has been slightly modified for print.}
%Yarvin2006a, http://lambda-the-ultimate.org/node/1269
%Yarvin2006, https://web.archive.org/web/20060701070718/http://urbit.sourceforge.net/u.txt

Extensive commentary on the operators is provided.  Rightwards grouping of tuple expressions has already been introduced.  Extension of the language is summarily ruled out.\footnote{Compare Ax and Conk, pp.~XX–XX herein.}  Data are conceived of as Unix-like byte streams; details of parsing and lexing are considered.  Terms (the ancestor of nouns) include a \texttt{NULL}-like ``foo'' type \texttt{\textasciitilde} distinguishable by value rather than structure.  \textsc{ascii} is built in as numeric codes, similar to G\"{o}del numbering.

As commenter Mario B.\ pointed out, the \texttt{U} specification permits \textsc{ski} operators with the simple expressions,

@@

While early work (1940s--50s) had been carried out on ``minimal instruction set computers'' (\textsc{misc}s), it is more likely that Yarvin was influenced by contemporaneous work on ``reduced instruction set computers'' (\textsc{risc}s) in the 1980s and 90s.  Language proposals like that of \citeauthor{Madore2003}'s Unlambda and \citeauthor{Burger2006}'s Pico Lisp may have influenced Yarvin's design choices throughout this era.
%Madore2003 http://www.madore.org/~david/programs/unlambda/
%Burger2006 https://software-lab.de/radical.pdf

@@

\begin{lstlisting}[label=lst:u,caption={\texttt{U}, 31 January 2006.  The earliest extant patriarch of the Nock family.},style=listingcode]
U: Definition

@@

% In plain language, the rules of \texttt{U} are as follows:

% \begin{enumerate}[label=(\alph*)]
%   %a
%   \item  constant reduction
%   %b
%   \item  cell detection
%   %c
%   \item  atom detection
%   %d
%   \item  branch selection
%   %e
%   \item  branch selection
%   %f
%   \item  equality
%   %g
%   \item  increment
%   %h
%   \item
%   %i
%   \item
%   %j
%   \item
%   %k
%   \item
%   %l
%   \item  cell distribution
%   %m
%   \item  
%   %n
%   \item
%   %o
%   \item
%   %p
%   \item  hyper-Turing follow operator
%   %q
%   \item
%   %r
%   \item  evaluation of two (choke)
%   %s
%   \item  evaluation of one (choke)
% \end{enumerate}

Source:  \citet{Yarvin2006}

@@

Axiomatic operator \texttt{*} tar\footnote{We refer to Nock axiomatic operators via their modern aural \textsc{ascii} pronunciations.  While these evolved over time (to wit, \texttt{\textasciicircum} ``hat'' became ``ket''), to attempt to synchronize pronunciation with the era of a Nock release is a fool's errand.} is identified as a \texttt{GOTO}.\footnote{One can see the influence of this version's naming scheme on Atman's Ax, pp.~XX–XX herein.}

@@

Source:  \citet{Nock13K}

@@

Opcodes were reordered slightly.  Compound opcodes were introduced, such as a conditional branch and a static hint opcode.

@@

Source:  \citet{Nock12K}

@@

Opcodes were reordered slightly.  The conditional branch was moved to 2.  Composition, formerly at 2, was removed.

@@

Source:  \citet{Nock11K}

@@

Parentheses were replaced by brackets.  Opcodes were reordered slightly.  Hint syntax was removed.  Functionally, 11K and 10K appear very similar, particularly if the Watt (proto-Hoon) compiler is set up to produce variable declarations and compositions as the compound opcodes had them.

@@

Source:  \citet{Nock10K}

@@

The cell detection axiomatic operator underlying opcode 4 (cell detection) was changed from \texttt{\&} pam to \texttt{?} wut.  Versus 10K, 9K elides operator names in favor of definitions.  Other differences are likewise primarily terminological, such as the replacement of \texttt{Deep} \texttt{\&} pam with \texttt{?} wut.

This version of Nock was published on the Moron Lab blog in 2010 \citep{Yarvin2010c} as ``Maxwell's equations of software''.  Yarvin emphasized that Nock was intended to serve as ``foundational system software rather than foundational metamathematics'' (ibid.).  Yarvin also publicly expounded on the practicality of building a higher-level language on top of Nock at this point (ibid.):

\begin{quote}
To define a language with Nock, construct two nouns, \texttt{q} and \texttt{r}, such that \texttt{*[q r]} equals \texttt{r}, and \texttt{*[s *[p r]]} is a useful functional language. In this description,
\begin{itemize}
  \item  \texttt{p} is the function source;
  \item  \texttt{q} is your language definition, as source;
  \item  \texttt{r} is your language definition, as data;
  \item  \texttt{s} is the input data.
\end{itemize}
\end{quote}

\noindent
More concretely, Watt (the predecessor to Hoon) is defined as:

@@

\noindent
This remains the essential pattern followed to this day by higher-level languages targeting Nock as an \textsc{isa}.

Yarvin had prepared to virtualize Nock interpretation to expose a broader namespace for interaction with values than the ``strict'' subject of a formula \citep{Yarvin2010c}.

@@

Source:  \citet{Nock9K}

@@

The compound opcodes reappeared.  Opcode 6 defined a conditional branch.  Opcode 7 was described as a function composition operator.  Opcode 8 served to define variables.  Opcode 9 defined a calling convention.  The remaining opcodes are hints, but each serving a different purpose:

\begin{enumerate}
  \setcounter{enumi}{10}
  \item  consolidate for reference equality.
  \item  yield an arbitrary, unspecified hint.
  \item  label for acceleration (jet).
\end{enumerate}

Nock 8K received an uncharacteristic amount of commentary, given a preprint document prepared for presentation at the 42nd \textsc{iscie} International Symposium on Stochastic Systems Theory and Its Applications (\textsc{sss'10}) \citep{Yarvin2010b}.

\sloppy
Lambda was highlighted as a design pattern (a ``gate'' or stored procedure call) enabled by the ``core'' convention.  \mbox{Notably,} \lstinline[style=listingcode]{[[sample context] battery]} occurred in a different order than has been conventional since 2013 (emphasizing that the ubiquitous core pattern is a convention rather than a requirement).  Watt was revealed to have a different \textsc{ascii} pronunciation convention than Nock at this stage.

@@

Source:  \citet{Nock8K}

@@

@@

The three hint opcodes were refactored into two, a static and a dynamic hint, both at 10.

@@

Source:  \citet{Nock7K}

@@

The axiomatic operator for increment was changed from \texttt{\textasciicircum} ket to \texttt{+} lus.  Compound opcode syntax was reworked slightly.

@@

Source:  \citet{Nock6K}

@@

Compound opcode syntax was reworked slightly.  All trivial reductions of axiomatic operators were removed to the preface of the specification.

(For instance, a trivial ``cosmetic'' change was made to 5K's specification after it was publicly posted in order to synchronize it with the VM's behavior (\href{https://github.com/urbit/archaeology/commit/dd779c1e92d7200c2ce6c922052f1eb1d1642e4c}{\texttt{dd779c1}}).)

@@

Source:  \citet{Nock5K}

@@

The primary change motivating 5K to 4K was the introduction of an edit operator \texttt{\#} hax, which ameliorated the proliferation of cells in the Nock runtime's memory.\footnote{The date must be earlier than 27 September 2018; cf.\ \href{https://github.com/urbit/urbit/pull/1027}{\texttt{urbit/urbit} \#1027}.}  The edit operator is an optimization which makes modifications to a Nock data structure more efficient.  It's a notable example of a change motivated by the pragmatics of the runtime rather than theoretical or higher-level language concerns.\footnote{See \patp{niblyx-malnus}, pp.~XX–XX, this volume, for a verbose derivation of the edit operator and opcode 10 from the primitive opcodes.}

Opcode 5 (equality) was rewritten to more explicit with application of the cell distribution rule.  Opcodes 6–9 were rewritten to utilize the \texttt{*} tar operator rather than routing via opcode 2.  Opcode 11 (formerly opcode 10) was likewise massaged.  In general, preferring to express rules using \texttt{*} tar proved to be slightly more terse than utilizing opcode 2.

\begin{lstlisting}[label=lst:nock4k,caption={Nock 4K, \textit{terminus ad quem} 27 September 2018.},style=listingcode]
Nock 4K

@@

Source:  \citet{Nock4K}

@@

While deviations from the trunk line of the Nock family have been propsed at various points,\footnote{Notably, Ax (see pp.~XX–XX, this volume), \textsc{skew}, and \textsc{plan} (see pp.~XX–XX, this volume).} Nock itself has remained the definitional substrate of Urbit since its inception.  It has also been adopted as the primary \textsc{isa} of Nockchain and the NockApp ecosystem.

Why, then, do we contemplate further changes?  The \textsc{skew} proposal by \patp{siprel} and \patp{little-ponnys} argued that Nock 4K represented an undesirable saddle point in the design space of possible Nocks, itself a ``ball of mud'' \citep{Skew2020}.  While \textsc{skew} itself was not adopted, it inspired the development of Plunder and \textsc{plan} as a solid-state computing architecture sharing some ambitions with Urbit and Nock \citep{Plunder2025}.  A rigorously æsthetic argument can thus be sustained that Nock is not yet ``close enough'' to its final, diamond-perfect form to be a viable candidate.

While some have found this argument compelling, Urbit's core developers have elected to maintain work in the ``main line'' of traditional Nock as the system's target \textsc{isa}.  The Nock 4K specification is a good candidate, in this sense, for a ``final'' version of Nock, as it has been successfully used in production for several years.  It seems more likely that subsequent changes to Nock will derive not from alternative representations but from either dramatically more elegant expressions (e.g., of \mbox{opcode} 6 or a combinator refactor) or from an \mbox{implicit} under\-specification in the current Nock 4K which should be made \mbox{explicit}.

@@

Nock began life as a hyper-Turing machine language, a theoretical construct for the purpose of defining higher-level programming languages with appropriate affordances and semantics.  While its opcodes and syntax have gradually evolved over the course of two decades, the ambition to uproot the Unix ``ball of mud'' and replace it with a simple operating function amenable to reason has remained the north star of Urbit and Nock.  The history of Nock serves as an index of refinement as Yarvin and contributors sought to balance conciseness, efficiency, and practicality.
